Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-25 linear_adia.tex
minor edits to sec 2, 3, 4

\subsection{The effect of introducing a small but finite
  thermal relaxation time}\label{relax_pert}
We are interested in finite thermal relaxation
timescales $\beta > 0$, but it is instructive to
first ask  the more analytically tractable question: how 
do the eigenfrequencies and eigenfunctions change when we change
$\beta$ from zero to a small but finite value? For 
sufficiently small $\beta$ we expect the solution to only differ
slightly from a case with $\beta\equiv 0$. We thus perturb a solution
for $\beta=0$ to see the effect of thermal relaxation.

For definiteness, let us consider the simple solution 
\begin{align}
  \beta\equiv 0, \quad \delta v_z = 1,\quad \hat{\upsilon}^2 = \frac{1 +
  \ii h q \hat{k}}{1+\hat{k}^2}, \label{pert_basic} 
\end{align}
which solves Eq. \ref{nearly_iso_explicit} since $\delta
v_z=$constant and $\chi=1$ so that $B=C=0$. 
This is the lowest order VSI mode, the `fundamental corrugation mode',
where the vertical velocity is constant throughout the disk
column. Hereafter we shall simply refer to them as the fundamental
mode.

@@

A finite thermal relaxation time allow buoyancy
forces to stabilize vertical motions in sub-adiabatically stratified
disks. The $\zhat^2$ dependence in $\delta v_{z1}$ makes  
sense because it becomes significant at large $|\zhat|$, i.e. away from
the midplane where the effect of buoyancy first appears as $\beta$ is
increased from zero (since $N_z^2\propto z^2$ for a thin disk, see
Eq. \ref{nzsq_def}).

>>>>

\subsection{The effect of introducing a small but finite
  thermal relaxation time}\label{relax_pert}
We are interested in finite thermal relaxation
timescales $\beta > 0$, but it is instructive to
first ask  the more analytically tractable question: how 
do the eigenfrequencies and eigenfunctions change when we change
$\beta$ from zero to a small but finite value? For 
sufficiently small $\beta$ we expect the solution to only differ
slightly from a case with $\beta\equiv 0$. We thus perturb a solution
for $\beta=0$ to see the effect of finite cooling.

For definiteness, let us consider the simple solution 
\begin{align}
  \beta\equiv 0, \quad \delta v_z = 1,\quad \hat{\upsilon}^2 = \frac{1 +
  \ii h q \hat{k}}{1+\hat{k}^2}, \label{pert_basic} 
\end{align}
which solves Eq. \ref{nearly_iso_explicit} since $\delta
v_z=$constant and $\chi=1$ so that $B=C=0$. 
This is the lowest order VSI mode, the `fundamental corrugation mode',
where the vertical velocity is constant throughout the disk
column. Hereafter we shall simply call it the fundamental
mode.

@@

A finite cooling time allow buoyancy
forces to stabilize vertical motions in sub-adiabatically stratified
disks. The $\zhat^2$ dependence in $\delta v_{z1}$ makes  
sense because it becomes significant at large $|\zhat|$, i.e. away from
the midplane where the effect of buoyancy first appears as $\beta$ is
increased from zero (since $N_z^2\propto z^2$ for a thin disk, see
Eq. \ref{nzsq_def}).

### 2015-04-25 linear_iso.tex
minor notation corrections
minor edits to sec 2, 3, 4

\item We consider eigenvalues such that
  $|\upsilon^2|\ll \kappa^2,\,\Omega^2$; which correspond to inertial
  waves \citepalias{barker15}. We call this the low frequency  
  approximation.

\item We are mainly interested in the effect of vertical shear, 
%  i.e. $q\neq
%  0$ for vertically isothermal disks. 
  which is already captured
  explicitly in the linearized equations through the term
  $d\Omega/dz$. We thus ignore the vertical dependence of $\Omega$ and
  $\kappa$ elsewhere, and replace them with
  $\OmK$.  We call this the Keplerian approximation.

@@

 In terms of the dimensionless variables
\begin{align}
  \hat{z} = z/H,\quad \hat{k}=k_xH, \quad \hat{\upsilon} =\hat{\omega} +
  \ii\hat{\sigma}= \upsilon/\OmK,
\end{align}
where $\hat{\omega}$ and $\hat{\sigma}$ are real, the above
approximations lead to a single second order ordinary differential
equation, 
% Eq. \ref{vertiso_gov} becomes 
% \begin{align}
%   0=& \delta v_z ^{\prime\prime} + \left(1 + \ii h
%     q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime \notag\\
%   &+
%   \left[\hat{\upsilon}^2\left(\hat{k}^2+\chi\right) +
%     \left(  \chi + \ii h q\hat{k}\right)\ln\rho^{\prime\prime}\right.
%   \notag\\
%   &\left.- \ln\rho^{\prime
%       2}\left(\khat^2 -
%       \ii h
%       q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z,\label{vertiso_gov_nondim}
% \end{align}
% where $^\prime$ denotes $d/d\zhat$.
%\subsection{Thin-disk approximation}\label{analytic_relax}
%The governing equation, Eq. \ref{vertiso_gov_nondim}, is then   
\begin{align}
  \delta v_z^{\prime\prime} - z A\delta v_z^\prime +
  (B - C\zhat^2)\delta v_z = 0,\label{nearly_iso_explicit}
\end{align}
where $^\prime$ denotes $d/d\zhat$ and 
\begin{align}
  &A \equiv 1 + \ii h q \hat{k},\\
  &B \equiv \hat{\upsilon}^2\left(\chi + \hat{k}^2\right) -
  \left(\chi + \ii h q \hat{k}\right),\\
  &C \equiv \left(1-\chi\right)\left(\hat{k}^2 - \ii
    h q\hat{k}\right), 
\end{align}
where
\begin{align}
\chi = \frac{1-\ii\hat{\upsilon}\beta}{1-\ii\hat{\upsilon}\beta\gamma}.
\end{align}
The derivation of Eq. \ref{nearly_iso_explicit}  is outlined in
Appendix \ref{adia_improve}.

>>>>

\item We consider inertial waves with eigenfrequencies such that
  $|\upsilon^2|\ll \kappa^2,\,\Omega^2$ \citepalias{barker15}. We call
  this the low-frequency approximation.

\item We are mainly interested in the effect of vertical shear, 
%  i.e. $q\neq
%  0$ for vertically isothermal disks. 
  which is already captured
  explicitly in the linearized equations through the term
  $\p_z\Omega$. We thus ignore the vertical dependence of $\Omega$ and
  $\kappa$ elsewhere, and replace them with
  $\OmK$.  We call this the Keplerian approximation.

@@

 In terms of the dimensionless variables
\begin{align}
  \hat{z} = z/H,\quad \hat{k}=k_xH, \quad \hat{\upsilon} =\hat{\omega} +
  \ii\hat{\sigma}= \upsilon/\OmK,
\end{align}
where $\hat{\omega}$ and $\hat{\sigma}$ are real, the above
approximations lead to a single second order ODE, 
% Eq. \ref{vertiso_gov} becomes 
% \begin{align}
%   0=& \delta v_z ^{\prime\prime} + \left(1 + \ii h
%     q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime \notag\\
%   &+
%   \left[\hat{\upsilon}^2\left(\hat{k}^2+\chi\right) +
%     \left(  \chi + \ii h q\hat{k}\right)\ln\rho^{\prime\prime}\right.
%   \notag\\
%   &\left.- \ln\rho^{\prime
%       2}\left(\khat^2 -
%       \ii h
%       q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z,\label{vertiso_gov_nondim}
% \end{align}
% where $^\prime$ denotes $d/d\zhat$.
%\subsection{Thin-disk approximation}\label{analytic_relax}
%The governing equation, Eq. \ref{vertiso_gov_nondim}, is then   
\begin{align}
  \delta v_z^{\prime\prime} - z A\delta v_z^\prime +
  (B - C\zhat^2)\delta v_z = 0,\label{nearly_iso_explicit}
\end{align}
where $^\prime$ denotes $d/d\zhat$ and 
\begin{align}
  &A \equiv 1 + \ii h q \hat{k},\\
  &B \equiv \hat{\upsilon}^2\left(\chi + \hat{k}^2\right) -
  \left(\chi + \ii h q \hat{k}\right),\\
  &C \equiv \left(1-\chi\right)\left(\hat{k}^2 - \ii
    h q\hat{k}\right), 
\end{align}
where
\begin{align}
\chi = \frac{1-\ii\hat{\upsilon}\beta}{1-\ii\hat{\upsilon}\beta\gamma}.
\end{align}
The derivation of Eq. \ref{nearly_iso_explicit}  is outlined in
Appendix \ref{adia_improve}.

### 2015-04-25 paper.tex
minor edits to sec 2, 3, 4

\section{Thermodynamic requirement for the vertical shear instability}\label{vsi_require}     
In a stably stratified thin disk, the VSI  
requires a thermal timescale $t_c$ much shorter than the
disk dynamical timescale, 
\begin{align}
  t _c \ll \OmK^{-1}. 
\end{align}
where $\OmK$ is the Keplerian frequency (N13).    
This requirement -- that the cooling time be much shorter than an orbital period, 
which in turn is much smaller than the relevant oscillation period of vertically elongated gravity waves  -- is quite stringent.
It highlights the fact that vertical buoyancy is a very strong stabilizing influence.
Very strong radiative damping is thus required to weaken buoyancy and allow the weak vertical shear to drive instability.

To roughly quantify the required smallness of $t_c$, we consider a
vertically isothermal disk with aspect-ratio $ h$, 
 radial temperature profile $T \propto r^q$ and adiabatic index
$\gamma$. For a PPD, $ h \sim 0.05$, $q\sim -0.5$ and $\gamma\simeq 1.4$.

\begin{figure*}
  \includegraphics[width=\linewidth]{figures/rhoshearNz}
  \caption{The radial and vertical structure of a ``minimum mass" PPD model showing gas density (left, relative to the midplane density at 1 AU), vertical shear rate (middle) and
    buoyancy frequency (right).
    \label{eqm_structure} 
  }
\end{figure*}

For a thin disk, the vertical shear increases with height, $z$, from the midplane as 
\begin{align}\label{vshear_thin}
  r \frac{\p \Omega }{\p z} \simeq  h q \OmK \left(\frac{z}{2 H}\right),
\end{align}
for several disk scale heights, $H = hr$. 
%As noted in
%\S\ref{intro} this vertical variation is a source of free energy and
%thus vertical shear ($\p_z\Omega \neq0$) is destabilizing. 
%
%For $\gamma>1$ the disk is stably stratified with a real 
This destabilizing shear completes with the stabilizing vertical
buoyancy frequency $N_z$.  In a thin disk,     
\begin{align}\label{nz_thin}
  N_z^2 \simeq \left(\frac{\gamma-1}{\gamma}\right) \left(\frac{z}{H}\right)^2
  \OmK^2.  
\end{align}
Fig. \ref{eqm_structure} maps shear and buoyancy (as well as the gas density) 
in a fiducial PPD model, whose parameters are in Appendix \ref{mmsn} . %Here $\rho_0$
%is the midplane density and vertical shear rate and buoyancy
%frequencies are normalized by the local Keplerian frequency. 
Vertical shear is 
generally weak compared to buoyancy, suggesting stability.

Without any cooling, the Solberg-Hoiland criteria confirms that vertical buoyancy is strong
enough to ensure (axisymmetric) stability, see \S\ref{solberg}.  With radiative cooling, 
thermal fluctuations decay which, combined with pressure equilibration, reduces the effective buoyancy.

@@

How short must $t_c$ be for vertical shear to prevail?  We start by 
considering the Richardson number $\mathrm{Ri} = N_z^2/(r\p_z\Omega)^2$, a 
ratio of buoyant to shear energies.  Though not precisely applicable to our problem,
non-rotating shear flows are stable if $\mathrm{Ri} > 1/4$ (\citealp{chandrasekhar61}, 
also see \citealp{ys02, lee10} for applications to thin dust layers in PPDs).
Following \cite{urpin03} and \cite{townsend58},  we reduce the buoyant energy by the 
ratio of cooling to forcing timescales, $t_c |r \p_z\Omega| \leq 1$ (with no reduction in buoyant energy 
expected when this inequality fails).  With this reduction, the Richardson like criterion for 
shear \emph{instability} becomes
\begin{align}\label{bcrit_est}
  t_c \lesssim \frac{\left|r\p\Omega/\p z\right|} {N_z^2}. 
\end{align}

This result is far from rigorous, but it contains some useful insights.
 If we interpret Eq.\ \ref{bcrit_est} as a local criterion, then we see that 
 the minimum cooling time decreases with height as $t_c \propto 1/z$.
 This finding is relevant to the radiative damping of so called ``surface 
 modes," which we consider in \S\ref{numerical}.

 To roughly obtain a global cooling time criterion, we evaluate Eq.\ \ref{bcrit_est}
 at $z \sim H$.  To get the precise form we want, we choose $z = \gamma H / 2$ to 
 obtain the cooling requirement
 \begin{align}\label{prelim_bcrit}
 t_c <  h |q| \OmK^{-1}/ (\gamma - 1)
\end{align}   
for the VSI in a vertically isothermal, thin disk.  This estimate agrees with 
our central result, the cooling criterion that we derive more rigorously in \S\ref{iso_vsi_beta_crit}.

@@

We now consider whether, and where, thermal timescales in realistic PPDs
are short enough to support VSI growth.  Fig. \ref{bcrit_mmsn2d} plots the
 thermal timescales in our fiducial PPD model with a small grain dust opacity. 
 Since cooling by radiative diffusion depends on a perturbation's lengthscale, we
 plot $t_c$ for two values of the radial wavenumber,  $k_x$,  the relevant 
 component for VSI modes that are radially narrow.   The red line
marks the boundary implied by Eq. \ref{prelim_bcrit}.  We caution that 
 Eq. \ref{prelim_bcrit} applies when $t_c$ is vertically uniform.  This section is 
 aimed at basic insights, while \S\ref{application}
 analyzes the VSI with realistic spatial variations in $t_c$.

Due to the smaller values of $t_c$ in the outer disk, we generally expect the VSI to 
operate outside $\sim 10$ AU.  This conclusion changes however, if the dust opacity is reduced, 
slowing down cooling, as we show later (e.g.\ Fig.\ \ref{mmsn_bcrit_bcool}).

In the inner disk we see that thermal times are long in the midplane, but short at higher altitutdes. 
Again, this structure necessitates the more detailed analyses that follow.  Nevertheless, we can already 
see that due to radiative diffusion higher wavenumbers can cool more rapidly.  Thus smaller scale disturbances 
may trigger VSI at smaller disk radii.

However we cannot and do not find that VSI will 
always be relevant for sufficiently small perturbations.  For starters, very small wavelengths are in the optically 
thin cooling regime,  where cooling is independent of lengthscale.  Moreover, small scale perturbations 
are prone to viscous damping, and should drive weaker transport.

>>>>

\section{Thermodynamic requirement for the vertical shear instability}\label{vsi_require}     
In a stably stratified thin disk, the VSI  
requires a thermal timescale $t_c$ much shorter than the
disk dynamical timescale, 
\begin{align}
  t _c \ll \OmK^{-1}. 
\end{align}
where $\OmK$ is the Keplerian frequency (N13).    
This requirement --- that the cooling time be much shorter than an orbital period, 
which in turn is much smaller than the relevant oscillation period of vertically elongated gravity waves  --- is quite stringent.
It highlights the fact that vertical buoyancy is strongly 
stabilizing. Rapid radiative damping is thus required to weaken
buoyancy and allow the weak vertical shear to drive instability.

To roughly quantify the required smallness of $t_c$, we consider a
vertically isothermal disk with aspect-ratio $ h$, 
 radial temperature profile $T \propto r^q$ and adiabatic index
$\gamma>1$ so the disk is stably stratified.  For a PPD, $ h \sim 0.05$,
$q\sim -0.5$ and $\gamma\simeq 1.4$.

\begin{figure*}
  \includegraphics[width=\linewidth]{figures/rhoshearNz}
  \caption{The radial and vertical structure of a `minimum mass' PPD model showing gas density (left, relative to the midplane density at 1 AU), vertical shear rate (middle) and
    buoyancy frequency (right).
    \label{eqm_structure} 
  }
\end{figure*}

For a thin disk ($h\ll 1$), the vertical shear rate varies with height, $z$, from the midplane as 
\begin{align}\label{vshear_thin}
  r \frac{\p \Omega }{\p z} \simeq  h q \OmK \left(\frac{z}{2 H}\right),
\end{align}
where $H=hr$ is the characteristic disk scale height. 
%for several disk scale heights, $H = hr$. 
%As noted in
%\S\ref{intro} this vertical variation is a source of free energy and
%thus vertical shear ($\p_z\Omega \neq0$) is destabilizing. 
%
%For $\gamma>1$ the disk is stably stratified with a real 
This destabilizing shear competes with the stabilizing vertical
buoyancy frequency $N_z$.  In a thin disk,     
\begin{align}\label{nz_thin}
  N_z^2 \simeq \left(\frac{\gamma-1}{\gamma}\right) \left(\frac{z}{H}\right)^2
  \OmK^2.  
\end{align}
Fig. \ref{eqm_structure} maps the vertical shear rate and buoyancy
frequency (as well as the gas density) 
in a fiducial PPD model considered in \S\ref{application}. 
%, whose parameters are in Appendix \ref{mmsn} . %Here $\rho_0$
%is the midplane density and vertical shear rate and buoyancy
%frequencies are normalized by the local Keplerian frequency. 
Vertical shear is 
generally weak compared to buoyancy, suggesting stability. In fact, 
without any cooling, the Solberg-Hoiland criteria \citep{tassoul78}  
confirms that vertical buoyancy is strong
enough to ensure (axisymmetric) stability, see \S\ref{solberg}.  With radiative cooling, 
thermal fluctuations decay which, combined with pressure equilibration, reduces the effective buoyancy.

@@

How short must $t_c$ be for vertical shear to prevail?  We start by 
considering the Richardson number $\mathrm{Ri} = N_z^2/(r\p_z\Omega)^2$, a 
ratio of buoyant to shear energies.  Though not precisely applicable to our problem,
non-rotating shear flows are stable if $\mathrm{Ri} > 1/4$ (\citealp{chandrasekhar61}, 
also see \citealp{ys02, lee10} for applications to thin dust layers in PPDs).
Following \cite{urpin03} and \cite{townsend58},  we reduce the buoyant energy by the 
ratio of cooling to forcing timescales, $t_c |r \p_z\Omega| \leq 1$ (with no reduction in buoyant energy 
expected when this inequality fails).  With this reduction, the Richardson-like criterion for 
shear \emph{instability} becomes
\begin{align}\label{bcrit_est}
  t_c \lesssim \frac{\left|r\p\Omega/\p z\right|} {N_z^2}, 
\end{align}
which is Eq.  47 of \cite{urpin03}. %This result is far from rigorous, but it contains some useful insights.
If we interpret Eq.\ \ref{bcrit_est} as a local criterion, then we see that 
the maximum cooling time which permits instability decreases with height as $1/|z|$;  
the stabilizing effect of vertical buoyancy increases more rapidly away from
the midplane than the destabilizing effect of vertical shear.  
This finding is relevant to the radiative damping of so called `surface 
modes', which we consider in \S\ref{numerical}.

For vertically global disturbances considered in this work, 
% To roughly obtain a global cooling time criterion, we evaluate Eq.\ \ref{bcrit_est}
% at $z \sim H$.  To get the precise form we want, we choose $z = \gamma H / 2$ to 
we obtain the cooling requirement 
\begin{align}\label{prelim_bcrit}
  t_c <  h |q| \OmK^{-1}/ (\gamma - 1)
\end{align}   
for the VSI to operate effectively in a vertically isothermal, thin
disk. We derive this result more formally in
\S\ref{iso_vsi_beta_crit}, but notice that Eq. \ref{prelim_bcrit}
may be obtained from Eq. \ref{bcrit_est} by evaulating it at the
height $|z|=\gamma H/2$. %i.e., at $z=O(H)$

%This estimate agrees with 
%our central result, the cooling criterion that we derive more rigorously in \S\ref{iso_vsi_beta_crit}.

@@

We now consider whether, and where, thermal timescales in realistic PPDs
are short enough to support VSI growth.  Fig. \ref{bcrit_mmsn2d} plots the
 thermal timescales in our fiducial PPD model with a small grain dust opacity. 
 Since cooling by radiative diffusion depends on a perturbation's lengthscale, we
 plot $t_c$ for two values of the radial wavenumber,  $k_x$,  the relevant 
 component for VSI modes that are radially narrow.  The red line
marks the boundary implied by Eq. \ref{prelim_bcrit}.  We caution that 
 Eq. \ref{prelim_bcrit} applies when $t_c$ is vertically uniform.  This section is 
 aimed at basic insights; while more realistic spatial variations in
 $t_c$ is considered in \S\ref{application}. 
 %analyzes the VSI with realistic spatial variations in.

 Due to the smaller values of $t_c$ in the outer disk, we generally expect the VSI to 
 operate outside $\sim 10$ AU.  This conclusion changes however, if
 the dust opacity is reduced,  
slowing down cooling, as we show in \S\ref{application}. % (e.g.\ Fig.\
% \ref{mmsn_bcrit_bcool}). 
In the inner disk, thermal times are long in the midplane, but short at higher altitutdes. 
Again, this structure necessitates the more detailed analyses that follow.  Nevertheless, we can already 
see that due to radiative diffusion higher wavenumbers can cool more rapidly.  Thus smaller scale disturbances 
may trigger VSI at smaller disk radii.

On the other hand, one cannot obtain arbitrarily short cooling times
by invoking arbitrarily small scales (cf. \citealt{urpin03}, 
\citetalias{barker15}), because very small wavelengths are in the
optically thin cooling regime,  where cooling is independent of
lengthscale. Furthermore, while we do not include viscosity in our
idealized disk models, in reality small-scale perturbations will be 
prone to viscous damping, and should drive weaker transport.

% However, we cannot and do not find that VSI will always be relevant for
% sufficiently small perturbations because \begin{inparaenum}[(i)] \item very small wavelengths are in the optically 
% thin cooling regime,  where cooling is independent of lengthscale;
% and \item small scale perturbations 
% are more prone to viscous damping, and should drive weaker transport.
% \end{inparaenum}
